Chemical Tagging Using Minimum Message Length. I.\\ Gaussian Mixture Modelling

### 2017-04-19 articles/chemical-tagging/ms.tex
some words (probably bad)

\shorttitle{Chemical Tagging Using Minimum Message Length}
\shortauthors{Casey et al.}

@@

\title{Chemical Tagging Using Minimum Message Length}

@@

\begin{enumerate}
\item The number of clusters.
\item The number of latent factors (e.g., supernovae yields, AGB yields).
\item The multivariate factor loads of individual latent factors.
\item The relative weighting (yet another latent factor, which is however not
      common between all mixtures) of individual mixtures.
\item The relative \emph{scoring} of different latent factors on each star.
\item The intrinsic \emph{specific variances} of the individual abundances.
\item The multivariate mean abundances and covariance matrices of each star
      cluster, where the covariance matrix includes off-diagonal terms that
      describe the \emph{data} correlations (e.g., the correlations that result
      from trying to measure multiple abundances from the same data).
\item Atomic diffusion, veiling, and thermohaline mixing: The surface chemical
      abundances of stars \emph{will change} over its lifetime.
\item Probabilistic priors (perhaps very hard priors) on joint responsibilities
      of stars that are specified from astrometric constraints.
\item There are missing data. Accept it.
\item There are \emph{upper limits} on some data.\footnote{
        Spectroscopists are probably the worst in the Universe when it comes to
        reporting upper limits. They will report an `\emph{upper limit}', when
        what might mean is the \emph{absolute limit} where there is zero
        likelihood that the true value is above the given point estimate (which might as well be $+\infinity$, or
        they may describe a \emph{pseudo-1$\sigma$} limit which is estimated
        from the locally (yet covariant!) heteroskedastic noise in the data (which often might as well be $\approx$0).
        In reality, an upper limit ought to be given as a probability density
        function that is skewed in one direction, perhaps similar to the shape
        of a log-normal distribution. In any case, the confidence to which they
        describe their `\emph{upper limit}' ought to be given.}
\item There are correlations in the measurements that result from trying to
      measure two (or more) abundances from the same data. As an example,
      consider a rare -- yet important -- chemical abundance that has a single
      atomic absorption line which is blended with a more common elemental
      abundance. In all likelihood, there will be a strong positive correlation
      between the abundance derived between the rare and more common elemental
      abundance.
\end{enumerate}

@@

There we introduce the concept of MML,

@@

Here we introduce a simple probablistic approach to chemical tagging, and in
later sections we will formulate the objective function for MML. Our approach is
the simplistic in the sense that we will ignore many of the aforementioned
problems that will plague chemical tagging at scale. Specifically we make the
following assumptions:

\begin{itemize}
\item We assume that the data (e.g., the detailed chemical abundances of stars)
      can be represented by a mixture of multivariate Gaussian distributions.
\item We will assume that the number of \emph{true} multivariate Gaussian
      distributions is not known. This assumption applies to our experiements
      involving generated data, and those using real data.
\item In addition to not knowing the \emph{number} of multivariate normal
      distributions, we further assume that the means and covariance matrices of
      each mixture is unknown.
\item Here we assume that the covariance matrix of individual components can be
      described as `free` or `full', in the sense that the data within a cluster
      can have off-diagonal correlation terms, which are unknown.
\item In all experiments we will assume that the data have homoskedastic noise
      properties.
\item The relative weights (or mixing proportions) of different distributions
      is unknown. We further will assume a relative (or conditional)
      probability distribution for each star belonging to a given cluster. That
      is to say, we do not adopt a `\emph{hard selection}' approach to cluster
      modelling, where stars would be assigned as definitively belonging to one
      cluster or another.
      % TODO: should we actually introduce the partial assignment later when we 
      % talk about the Gaussian distributions?
\end{itemize}

We have made relatively few explicit assumptions about the model for the data,
but even the model described constitutes a very formidable model selection 
problem. The reason is because we do not know \emph{how many} multivariate
Gaussian distributions we ought to chose to fit the data. Including more
mixtures will consistently provide a better fit to the data (in a likelihood
sense), but we must make some decision or heuristic as to whether the additional
mixture is justified.

@@

\noindent{}where $\bm{\mu}$ and $\bm{\Sigma}$ is the multivariate mean and 
covariance matrix, and $\bm{y}$ are the data.  When modelling a mixture of
Gaussians, there are two approaches in determining assignments (or
responsibilities) of each star to a given cluster. One approach is to include
a (discrete) model parameter for every data point, which represents the index
of the mixture that the observation is assigned to. The second approach is to 
use a weighted mixture. These two are mathematically identical.
% TODO

@@

We assume a uniform prior on the mixing weights $\bm{w}$, only requiring that
their summation equals one. We adopt a wide uniform prior of $\mathcal{U}(\bm{\mu}) = [-12, 12]$ on all
multivariate mean abundances $\bm{\mu}$ (in all $d$ dimensions). This prior
is sufficiently wide that it has no impact on our inferences (24 orders of
magnitude!), but remains proper. We adopt a congujate inverted Wishart prior 
for the covariance matrices in the individual mixtures. Ignoring the effect of 
our prior on $\bm{\mu}$ gives us a proportional joint prior,

\begin{equation}
    p(\bm{\mu},\bm{\Sigma}) \propto |C|^{\frac{d + 1}{2}} 
\end{equation}
% TODO: DERIVE THIS.

@@

\acknowledgments
Who?
Grants.

>>>>

\shorttitle{Chemical Tagging Using Minimum Message Length. I.}
\shortauthors{Casey et al.}

@@

\title{Chemical Tagging Using Minimum Message Length. I.\\
       Gaussian Mixture Modelling}

@@

\begin{enumerate}
\item The number of star clusters (or star-forming events) where at least one
      star was observed and detailed chemical abundances were derived.
\item The number of latent factors (e.g., supernovae yields, AGB yields).
      Within the chemical tagging literature, this is commonly referred to as
      the number of `effective chemical dimensions'.
\item The multivariate factor loads of individual latent factors. For example,
      if core-collapse supernovae and AGB stars had both contributed to the
      chemical abundances of all stars in the sample, then the
      \emph{factor loads} would be the \emph{yield} of the supernovae, or the
      yield of relative abundances that were produced by the AGB star. Each
      star will be affected differently by these yields (see next item).
\item The relative \emph{scoring} of different latent factors on each star.
      This is a scaling that is applied to the latent yields. Continuing our
      analogy on yields from core-collapse supernovae and AGB stars, the
      \emph{factor scores} can be thought of as the relative contributions of
      the different factor loads, for a single star.
\item The relative weighting (yet another latent factor, which is however not
      common between all mixtures) of individual mixtures.
\item The multivariate mean abundances and covariance matrices of each star
      cluster, where the covariance matrix includes off-diagonal terms that
      describe the \emph{data} correlations (e.g., the correlations that result
      from trying to measure multiple abundances from the same data).
\item The intrinsic \emph{specific variances} of the individual abundances.
\item Atomic diffusion, veiling, and thermohaline mixing: The surface chemical
      abundances of stars \emph{will change} over its lifetime.
\item Probabilistic priors (perhaps very hard priors) on joint responsibilities
      of stars that are specified from astrometric constraints.
\item There are missing data. 
\item There are \emph{upper limits} on some data.\footnote{
        Spectroscopists are probably the worst in the Universe when it comes to
        reporting upper limits. They will report an `\emph{upper limit}', when
        what might mean is the \emph{absolute limit} where there is zero
        likelihood that the true value is above the given point estimate (which might as well be $+\infinity$, or
        they may describe a \emph{pseudo-1$\sigma$} limit which is estimated
        from the locally (yet covariant!) heteroskedastic noise in the data (which often might as well be $\approx$0).
        In reality, an upper limit ought to be given as a probability density
        function that is skewed in one direction, perhaps similar to the shape
        of a log-normal distribution. In any case, the confidence to which they
        describe their `\emph{upper limit}' ought to be given.}
\item There are correlations in the measurements that result from trying to
      measure two (or more) abundances from the same data. As an example,
      consider a rare -- yet important -- chemical abundance that has a single
      atomic absorption line which is blended with a more common elemental
      abundance. In all likelihood, there will be a strong positive correlation
      between the abundance derived between the rare and more common elemental
      abundance.
\end{enumerate}

@@

In this \article\ we will introduce a simple probabilistic approach to
chemical tagging, where the data are modelled as a mixture of multivariate
Gaussian distributions (Section \ref{sec:the-model}). In Section \ref{sec:mml}
we introduce the principle of minimum message length (MML), and formulate our
objective function by minimising the total message length. Our model is
simplistic in the sense that we will ignore many of the self-consistency
issues already introduced. However, in Section \ref{sec:experiments} we show
that our MML-based approach is superior to all other methods considered for
chemical tagging, all of which cannot alleviate the self-consistency issues.
In Section \ref{sec:discussion} we will describe how MML can overcome these
aforementioned issues, in order to construct a fully-consistent probabilistic
approach to chemical tagging.

@@

We make the following explicit assumptions:

\begin{itemize}
\item We assume that the data (e.g., the $D$ detailed chemical abundances of 
      $N$ stars) can be represented by a mixture of $K$ multivariate Gaussian
      distributions.
\item We assume that there are no missing data values. For example, if a star
      has been observed, then there is a complete set of $D$ chemical 
      abundances available.
\item We will assume that the number of \emph{true} multivariate Gaussian
      distributions $K_{true}$ is not known. This assumption applies to our 
      experiements involving generated data, and those using real data.
\item In addition to not knowing the \emph{number} of multivariate normal
      distributions, we further assume that the means $\bm{\mu}$ and 
      covariance matrices $\bm{C}$ of each $K$th mixture is unknown.
\item We assume that the covariance matrix of individual components can be
      described as `free' or `full', insomuch that the data within a cluster
      can have off-diagonal correlation terms, and those terms are unknown.
\item In all experiments we will assume that the data have homoskedastic noise
      properties.
\item The relative weights $\bm{w}$ (or mixing proportions) of $K$ 
      distributions is unknown. 
\item We assume a relative (or conditional) probability distribution for each 
      star belonging to a given cluster. That is to say that we do not adopt a 
      `\emph{hard selection}' approach to cluster modelling, where stars would
      be assigned as definitively belonging to one cluster or another. Our
      approach could be described as `\emph{soft} clustering'. This assumption
      has practical numerical outcomes and implications for the message length.
\end{itemize}

We have made relatively few explicit assumptions about the model for the data,
but even the model described constitutes a very formidable model selection 
problem. This is because we do not know \emph{how many} multivariate Gaussian
distributions we ought to chose to fit the data. Including more mixtures will
consistently provide a better fit to the data (in a likelihood sense), but we 
must make some decision or heuristic as to whether the additional mixture is 
justified.

@@

\noindent{}where $\bm{\mu}$ and $\bm{\Sigma}$ is the multivariate mean and 
covariance matrix, and $\bm{y}$ are the data.

% Full probability for the data.

When modelling a mixture of
Gaussians, there are two approaches in determining assignments (or
responsibilities) of each star to a given cluster. One approach is to include
a (discrete) model parameter for every data point, which represents the index
of the mixture that the observation is assigned to. The second approach is to 
use a weighted mixture. These two are mathematically identical.
% TODO

@@

We assume a uniform prior on the mixing weights $\bm{w}$, only requiring that
$\sum_{k=1}^{K}w_k = 1$. A useful extension of this work might be to impose
a Dirchlet prior on the mixing weights $\bm{w}$, or a distribution inspired by
the stellar mass function and observed selection function.

We adopt an improper uniform prior of 
$\mathcal{U}(\bm{\mu}) = [-\infinity, \infinity]$ on all multivariate mean 
abundances $\bm{\mu}$ (in all $D$ dimensions). This prior has no impact on
our inferences and becomes proper when it enters the (accepted) posterior
distribution. We similarily could adopt a proper and large uniform prior on all 
mean abundances of $\mathcal{U}(\bm{\mu}) = [-12, 12]$ and the result would be
the same.

We adopt a congujate inverted Wishart prior for the covariance matrices in the
individual mixtures. The joint prior density for the parameters of a single
mixture is given by the limiting form of the normal inverted-Wishart
density \citep[e.g., Section 5.2.3 of ][]{Schafer_1990}:

\begin{equation}
    p(\bm{\mu},\bm{\Sigma}) \propto |C|^{\frac{d + 1}{2}} \quad .
\end{equation}

@@

\acknowledgments

This research is supported by an Australian Research Council Discovery Project
grant (DP160100637). 
Source code for this project is available at \texttt{\giturl}. This document
was compiled on \gitdate\ from revision hash \texttt{\githash} in that
repository.
